Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-31 sections/lang.tex
update
Update on Overleaf.

The syntax of Finch is displayed in Figure \ref{fig:syntax}. The Finch syntax
mirrors most imperative languages with for-loops and control flow. Notable
inclusions to the language at the level of statements includes \mintinline{julia}{for},
\mintinline{julia}{let}, blocks of code with \mintinline{julia}{if}, and 
the lifecycle functions that let us declare, freeze, and thaw tensors.
%
At the level of expressions, we support a wide array of scalar operations through literals and calls to functions whose properties and definitions are defined externally.
%
The expressions can also interact with indices and extents.
% FIXME: ???
Finally, as detailed in the previous section, tensors are defined externally via an interface that supports the declare, freeze, thaw, and unfurl functions.
%
The first three are supported directly in the syntax whereas the four will be introduced through evaluation of \mintinline{julia}{for}-loops loops and accesses,
in the next section.
%

Our syntax is highly permissive: by allowing blocks of code with multiple statements, we support both multiple outputs and temporary tensors, seen most prominently in our implementation of gustavson's algorithm for sparse-sparse matrix multiply and breadth-first-search.
%
Our breadth-first seach also uses an
\mintinline{julia}{if} statement to avoid operating on vertices outside the
frontier. 
%
We have found that \mintinline{julia}{if} is particularly useful in
our model for masking to avoid work.
%
However, to compile these constructs efficient, we must restrict this syntax in several ways, as we have alluded to in the previous section via life cycles.
%
In the semantics section, we detail other specifics that allow us to specialize our syntax efficient code over structured data.

@@

We include the sketch of our semantics to highlight several key design choices.
%
\teo{Make this true:}
First, we note that the life cycle rules (in combination with the definition rules) prevent programs that read and write to values within a loop except in the narrow circumstance of updates, which cannot even be applied to scalar definitions.
%
Beyond the simple management of iterator state, these rules allow us to effectively analyze the propagation of constants within a loop.
%
We eliminate problems that in a less restrictive language would require alias analysis so that we can tell when a zero or repetitive value read from one tensor is being used in an expression or written to another tensor.
%
These restrictions play a similar role to that of a pure function in Halide~\cite{ragan-kelley_halide_2013}.
%
Second, loops with tensor accesses enter into a looplet system via the Unfurl rule.
%
In this system, repeated structures (Run) and constant information (Switch)  are slowly uncovered as accesses are lowered in various points in the program.
%
In this process, we are able to use rewrite rules (Simplify) to eliminate cases, unnecessary iterations, and so forth based on the information provided via Looplets and via the control flow (loops, sieve, definitions).
%

Though the user may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language ultimately through the assurance that accesses to tensor values will first unravel to Looplets and then simplify in the context of the location of the access.
%
In the next section, we will handle complex index access expressions involving indirection and arithmetic by transforming complex accesses to additional loop variables with additional tensor accesses to insert Looplets to provide the structure dictated by the complex control flow.
%
Also in the next section, We will show how similar means lets us exploit the structure of complex conditional expressions when the structure of the expressions is placed inside a tensor access.
\teo{More next sections teaserse here}

@@

% \begin{multline*}
%         Modes:=update \mid read \mid undefined \mid defined
% \end{multline*}
    \begin{prooftree}
    \hypo{\langle val, (e, t) \rangle
    \rightarrow val'}
    \hypo{var\not\in e}
    \infer2[$Define$]{\splitfrac{\langle\finchdefine(var, val, body), (e, t)\rangle}{\rightarrow \langle body, (e[var \mapsto val'], t) \rangle}}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{\langle args_i, (e, t) \rangle \Rightarrow vals_i}
        \hypo{\langle f, (e, t) \rangle \Rightarrow g}
        \infer2[$Call$]{\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \llangle g(vals...), t \rrangle}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}
        \hypo{}
        \infer1[$Literal$]{\langle\finchliteral(val), (e, t)\rangle \rightarrow val}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{}
        \infer1[$Value$]{\langle\finchvalue(ex, type), (e, t)\rangle \Rightarrow \llangle ex, t \rrangle}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}
        \infer0[$Variable$]{\splitfrac{\langle\finchvar(name), (e, t)\rangle}{\rightarrow e(\finchvar(name))}}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \infer0[$Index$]{\splitfrac{\langle\finchindex(name), (e, t)\rangle}{\rightarrow e(\finchindex(name))}}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}
    \hypo{\langle head,s_{1} \rangle \rightarrow s_2}
    \infer1[$BlockHead$]{\langle\finchblock(head, tail...), s_1 \rangle \rightarrow \langle \finchblock(tail...), s_2 \rangle}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \infer0[$EmptyBlock$]{\langle\finchblock(), s\rangle \rightarrow s}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}
        \hypo{\langle cond, s \rangle \Rightarrow true}
        \infer1[$SieveTrue$]{\langle\finchsieve(cond, body), s\rangle \rightarrow \langle body, s \rangle}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{\langle cond, s \rangle \Rightarrow false}
        \infer1[$SieveFalse$]{\langle\finchsieve(cond, body), s\rangle \rightarrow s}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{\finchmode(tns)\not\in e}
    \hypo{e(tns) = tns'}
    \hypo{\langle init, s \rangle \Rightarrow init'}
    \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
    \infer5[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle declare(tns', init', dims'...), t \rrangle)}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(\finchmode(tns)) = update}
    \hypo{e(tns) = tns'}
    \infer3[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \llangle freeze(tns'), t \rrangle)}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(\finchmode(tns)) = read}
    \hypo{e(tns) = tns'}
    \infer3[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle thaw(tns'), t \rrangle)}
    \end{prooftree}

    \caption{Basic evaluation semantics, roughly defining most of these language
    constructs to function similarly to their classical definitions. Note that
    The state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
    environment and another state $t$ corresponding to the state in the
    host language. Several Looplets introduce variables into the embedding
    language, which may be read when evaluating the \finchvalue node.
    %
    The state $s$ of the finch compiler
    is a tuple $(e, t)$ of a variable value environment and another state $t$
    corresponding to the state in the host language.
    This means that rules which modify $t$ are running in the host language. All of the lifecycle
    functions are designed to be implemented and executed in the host language,
    but these semantics enforce that each of these functions may update state in
    the host language and flip the mode of the tensor between read and update}
    \label{fig:semantics_core}
\end{figure}

@@

    \caption{Looplet evaluation semantics. 
    The state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
    environment and another state $t$ corresponding to the state in the host
    language. This means that rules which modify $t$ are running in the host language. Note that $E$ is an evaluation context that applies anywhere in
    the syntax tree. The nonlocal evaluations of Looplets are what allow looplets to
    hoist conditions and subranges out of loops. However, this also means we must specify
    the priority in which we apply looplet rules, which is as follows:
    $Thunk > Phase > Switch > Simplify > Run > Spike > Sequence > StepperSeek > StepperNext > Lookup > Unfurl$
    Many looplets, most notably the thunk looplet, introduce variables into the
    host language environment.  While variables introduced by a looplet may be
    modified by the looplet itself (steppers often increment some state
    variables), we forbid child looplets from modifying any state variables that
    they didn't introduce. This allows us to treat the \finchvalue node as a
    constant.
    %
    Note: the $Simplify$ rule references $algebra$, which is our variable
    defining a set of straightforward simplification rules. These rules include
    simple properties like $x * 0 \rightarrow 0$ to more complicated ones such
    as constant propagation. We omit the full set of rules for brevity, but
    point the curious reader to \cite[Figure 5]{ahrens_looplets_2023} for some examples}
    \label{fig:semantics_looplets}
\end{figure}

>>>>

The syntax of Finch is displayed in Figure \ref{fig:syntax}. The Finch syntax
mirrors most imperative languages with for-loops and control flow. Notable
statements that have been added to the language include \mintinline{julia}{for},
\mintinline{julia}{let}, blocks of code with \mintinline{julia}{if}, and 
the lifecycle functions that let us declare, freeze, and thaw tensors.
%
At the level of expressions, we support a wide array of scalar operations through literals and calls to functions whose properties and definitions are defined externally.
%
The expressions can also interact with indices and extents.
% FIXME: ???
Finally, as detailed in the previous section, tensors are defined externally via an interface that supports the declare, freeze, thaw, and unfurl functions.
%
The first three are supported directly in the syntax whereas the four will be introduced through evaluation of \mintinline{julia}{for}-loops loops and accesses,
in the next section.
%

Our syntax is highly permissive: by allowing blocks of code with multiple statements, we support both multiple outputs and temporary tensors, seen most prominently in our implementation of gustavson's algorithm for sparse-sparse matrix multiply and breadth-first-search.
%
Our breadth-first seach also uses an
\mintinline{julia}{if} statement to avoid operating on vertices outside the
frontier. 
%
We have found that \mintinline{julia}{if} is particularly useful in
our model for masking to avoid work.
%
However, to compile these constructs efficiently, we must restrict this syntax in several ways, as we have alluded to in the previous section, via life cycles.
%
In the semantics section, we detail other specifics that allow us to specialize our syntax efficient code over structured data.

@@

We include the sketch of our semantics to highlight several key design choices.
%
First, we note that the life cycle rules (in combination with the definition rules) prevent programs that read and write to values within a loop except in the narrow circumstance of updates, which cannot even be applied to scalar definitions.
%
Beyond the simple management of iterator state, these rules allow us to effectively analyze the propagation of constants within a loop.
%
We eliminate problems that in a less restrictive language would require alias analysis so that we can tell when a zero or repetitive value read from one tensor is being used in an expression or written to another tensor.
%
These restrictions play a similar role to that of a pure function in Halide~\cite{ragan-kelley_halide_2013}.
%
Second, loops with tensor accesses enter into a looplet system via the Unfurl rule.
%
In this system, repeated structures (Run) and constant information (Switch)  are slowly uncovered as accesses are lowered in various points in the program.
%
In this process, we are able to use rewrite rules (Simplify) to eliminate cases, unnecessary iterations, and so forth based on the information provided via Looplets and via the control flow (loops, sieve, definitions).
%

Though the user may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language ultimately through the assurance that accesses to tensor values will first unravel to Looplets and then simplify in the context of the location of the access.
%
In the next section, we will handle complex index access expressions involving indirection and arithmetic by transforming complex accesses to additional loop variables with additional tensor accesses to insert Looplets to provide the structure dictated by the complex control flow.
%
Also in the next section, we will show how similar means lets us exploit the structure of complex conditional expressions when the structure of the expressions is placed inside a tensor access.

@@

% \begin{multline*}
%         Modes:=update \mid read \mid undefined \mid defined
% \end{multline*}
    \begin{prooftree}
    \hypo{\langle val, (e, t, d) \rangle
    \rightarrow val'}
    \hypo{var\not\in d}
    \infer2[$Define$]{\splitfrac{\langle\finchdefine(var, val, body), (e, t, d)\rangle}{\rightarrow \langle body, (e[var \mapsto val'], t, d\cup\{var\}) \rangle}}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{\langle args_i, (e, t) \rangle \Rightarrow vals_i}
        \hypo{\langle f, (e, t) \rangle \Rightarrow g}
        \infer2[$Call$]{\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \llangle g(vals...), t \rrangle}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}
        \hypo{}
        \infer1[$Literal$]{\langle\finchliteral(val), (e, t, d)\rangle \rightarrow val}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{}
        \infer1[$Value$]{\langle\finchvalue(ex, type), (e, t, d)\rangle \Rightarrow \llangle ex, t \rrangle}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}
        \infer0[$Variable$]{\splitfrac{\langle\finchvar(name), (e, t, d)\rangle}{\rightarrow e(\finchvar(name))}}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \infer0[$Index$]{\splitfrac{\langle\finchindex(name), (e, t, d)\rangle}{\rightarrow e(\finchindex(name))}}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}
    \hypo{s = (e, t, d)}
    \hypo{\langle head, (e, t, \{\}) \rangle \rightarrow s^{\prime}}
    \hypo{\langle \finchblock(tail...), s^{\prime} \rangle \rightarrow (e^{\prime}, t^{\prime}, d^{\prime})})
        \infer3[$BlockHead$]{\langle\finchblock(head, tail...), s \rangle \rightarrow (e^{\prime}, t^{\prime}, d) }
    \end{prooftree}
    \hfill

    \vspace{6pt}

    \begin{prooftree}
    \hypo{\langle cond, (e, t, d) \rangle \Rightarrow true}
    \hypo{\langle body, (e, t, \{\}) \rangle \rightarrow (e^{\prime}, t^{\prime}, d^{\prime})}
        \infer2[$SieveTrue$]{\langle\finchsieve(cond, body), (e, t, d)\rangle \rightarrow (e^{\prime}, t^{\prime}, d)}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{\langle cond, s \rangle \Rightarrow false}
        \infer1[$SieveFalse$]{\langle\finchsieve(cond, body), s\rangle \rightarrow s}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}  
    \hypo{s = (e, t, d)}
    \hypo{tns\not\in d}
    \hypo{e(tns) = tns'}
    \hypo{\langle init, s \rangle \Rightarrow init'}
    \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
    \infer5[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle declare(tns', init', dims'...), t \rrangle, d\cup\{ tns\})}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}  
    \hypo{s = (e, t,d )}
    \hypo{e(\finchmode(tns)) = update}
    \hypo{tns\in d}
    \hypo{e(tns) = tns'}
    \infer4[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \llangle freeze(tns'), t \rrangle, d)}
    \end{prooftree}
    \vspace{6pt}

    \begin{prooftree}  
    \hypo{s = (e, t, d)}
    \hypo{e(\finchmode(tns)) = read}
    \hypo{tns\in d}
    \hypo{e(tns) = tns'}
    \infer4[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle thaw(tns'), t \rrangle, d)}
    \end{prooftree}

    \caption{Basic evaluation semantics, roughly defining most of these language
    constructs to function similarly to their classical definitions.
    %
    The state $s$ of the finch compiler is a tuple $(e, t, d)$ of a variable value
    environment, another state $t$ corresponding to the state in the
    host language, and finally the set of tensors defined within the current scope, $d$. 
    %
    This means that rules which modify $t$ are running in the host language.
    %
    Several Looplets introduce variables into the embedding
    language, which may be read when evaluating the \finchvalue node.
%
     All of the lifecycle
    functions are designed to be implemented and executed in the host language,
    but these semantics enforce that each of these functions may update state in
    the host language and flip the mode of the tensor between read and update}
    \label{fig:semantics_core}
\end{figure}

@@

    \caption{Looplet evaluation semantics. 
    %
    Basic evaluation semantics, roughly defining most of these language
    constructs to function similarly to their classical definitions.
    %
    The state of the compiler is described in the previous figure.
    %
     Note that $E$ is an evaluation context that applies anywhere in
    the syntax tree. 
    %
    The nonlocal evaluations of Looplets are what allow looplets to
    hoist conditions and subranges out of loops.
    %
    However, this also means we must specify
    the priority in which we apply looplet rules, which is as follows:
    $Thunk > Phase > Switch > Simplify > Run > Spike > Sequence > StepperSeek > StepperNext > Lookup > Unfurl$
    %
    Many looplets, most notably the thunk looplet, introduce variables into the
    host language environment. 
    %
    While variables introduced by a looplet may be
    modified by the looplet itself (steppers often increment some state
    variables), we forbid child looplets from modifying any state variables that
    they didn't introduce. 
    %
    This allows us to treat the \finchvalue node as a
    constant.
    %
    Note: the $Simplify$ rule references $algebra$, which is our variable
    defining a set of straightforward simplification rules. 
    %
    These rules include
    simple properties like $x * 0 \rightarrow 0$ to more complicated ones such
    as constant propagation. 
    %
    We omit the full set of rules for brevity, but
    point the curious reader to \cite[Figure 5]{ahrens_looplets_2023} for some examples}
    \label{fig:semantics_looplets}
\end{figure}

### 2024-03-31 sections/tensors.tex
Update on Overleaf.

\section{Bridging Looplets and Finch: The Tensor Interface}
%
The Finch language provides descriptions of computations that iterate over a subset of a regular grid that is lexicographically ordered.
%
At this point, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators, but Finch programs and data structures are sufficiently flexible that this impossible.
%
First, the Finch language interacts with multi-dimensional tensors whereas the Looplet abstraction is best suited towards iterators over a single dimension.
%
We require a bridge between the single dimensional iterators created from looplets and the mutli-dimensional abstractions common to tensor compilers.
%
Second, since the iteration order of a Finch program might not match that of a data structure (a discordant traversal), different iterators need to be requested for the same data depending on the traversal order of the program.
%
So we require a bridge that can provide different iteration orders depending on the context.
%
Third, since Finch programs can read and write to the same data, multi-dimensional tensors need to provide iterators for reading and writing as well as machinery to manage transition between these states.

@@

The first three of these functions, declare, freeze, and thaw, have to do with
managing when tensors can be assumed mutable or immutable. As we use Looplets to
represent iteration over a tensor, we must restrict the mutability of tensors
while we iterate over them. For example, if a tensor declares it has a constant
region from $i = 2:5$, but some other part of the computation modifies the
tensor at $i = 3$, this would result in incorrect behavior. It is much easier to
write correct Looplet code if we can assume that the tensor is immutable while
we are reading from it. Thus, we introduce the notion that a tensor can be in
read-only mode or update-only mode.  In read-only mode, the tensor may only
appear in the right-hand side of assignments. In update-only mode, the tensor
may only appear in the left-hand side of an assignment, either being overwritten
or incremented by some operator.  We can switch between these modes using freeze
and thaw functions. The declare function is used to allocate a tensor,
initialize it to some specified size and value, and leave it in update-only
mode.

@@

As future work, we also recommend the implementation of a tree-based format, as
this would enable incremental construction. Currently, the cost of freezing and
thawing is usually proportional to the number of stored fibers, so an operation
that sets a single value in a tensor would be prohibitively expensive (i.e.
$A[3, 4] = 3.14$). We would also recommend supporting SparseRLE as an
intermediate randomly accessible structure, because (unlike blocks or
singletons), runs have asymptotic benefits. However, these cases were not
necessary for our case studies.

>>>>

\section{Bridging Looplets and Finch: The Tensor Interface}
%
The Finch language provides descriptions of computations that iterate over a subset of a regular grid that is lexicographically ordered.
%
At this point, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators, but Finch programs and data structures are sufficiently flexible that this impossible\saman{?????}.
%
First, the Finch language interacts with multi-dimensional tensors whereas the Looplet abstraction is best suited towards iterators over a single dimension.
%
We require a bridge between the single dimensional iterators created from looplets and the mutli-dimensional abstractions common to tensor compilers.
%
Second, since the iteration order of a Finch program might not match that of a data structure (a discordant traversal), different iterators need to be requested for the same data depending on the traversal order of the program.
%
So we require a bridge that can provide different iteration orders depending on the context.
%
Third, since Finch programs can read and write to the same data, multi-dimensional tensors need to provide iterators for reading and writing as well as machinery to manage transition between these states.

@@

The first three of these functions, declare, freeze, and thaw, \saman{Italics for keywords?}have to do with
managing when tensors can be assumed mutable or immutable. \saman{This is because Finch goes beyond just a single einsum...the scope of Finch w.r.t. TACO etc, is not clear.} As we use Looplets to
represent iteration over a tensor, we must restrict the mutability of tensors
while we iterate over them. For example, if a tensor declares it has a constant
region from $i = 2:5$, but some other part of the computation modifies the
tensor at $i = 3$, this would result in incorrect behavior. It is much easier to
write correct Looplet code if we can assume that the tensor is immutable while
we are reading from it. Thus, we introduce the notion that a tensor can be in
read-only mode or update-only mode.  In read-only mode, the tensor may only
appear in the right-hand side of assignments. In update-only mode, the tensor
may only appear in the left-hand side of an assignment, either being overwritten
or incremented by some operator.  We can switch between these modes using freeze
and thaw functions. The declare function is used to allocate a tensor,
initialize it to some specified size and value, and leave it in update-only
mode.

@@

\saman{Can we not put future work into the submission!}
As future work, we also recommend the implementation of a tree-based format, as
this would enable incremental construction. Currently, the cost of freezing and
thawing is usually proportional to the number of stored fibers, so an operation
that sets a single value in a tensor would be prohibitively expensive (i.e.
$A[3, 4] = 3.14$). We would also recommend supporting SparseRLE as an
intermediate randomly accessible structure, because (unlike blocks or
singletons), runs have asymptotic benefits. However, these cases were not
necessary for our case studies.
